A Plasma Torus Around a Young Low-Mass Star

### 2025-01-19 paper/ms.tex
update

%\subsection{Introduction}
M dwarfs, stars with masses below about half that of the Sun, are the
only type of star to offer near-term prospects for detecting the
atmospheres of rocky exoplanets with water on their surfaces (CITE).
Investment with JWST has proceeded accordingly (CITE CITE).  It is
therefore important to consider how the evolution of an M dwarf might
impact the evolution of its planets.  Previous work has established
that most M dwarfs host close-in planets (CITE), and that these
planets are often subject to long circumstellar disk lifetimes (CITE),
to large doses of UV radiation (CITE), and to a high incidence of
flares and coronal mass ejections (CITE).  However, despite excellent
work in these areas, the properties of the circumstellar plasma
and magnetospheric environments to which young, close-in exoplanets
are subject remain largely unexplored.

One glaring example of our current ignorance is the complex periodic
variables (CPVs).  Figure~\ref{fig:lc} highlights the main object of
interest in this article, but over one hundred analogous objects have
now been discovered by K2 and TESS (CITE, CITE, CITE, CITE).  These CPVs are
defined by their highly structured and periodic optical light curves, 
and most are M dwarfs with rotation periods shorter than two days.
Within current sensitivity limits, none have primordial disks (CITE).
However, $\approx$3\% of stars a few million years old show this
behavior (CITE), and the observed fraction decreases to $\approx$0.3\%
by $\approx$150\,Myr (CITE).

The two leading hypotheses to explain the CPVs are either that
transiting clumps of circumstellar material corotate with the star
(CITE, CITE, CITE), or that these stars represent an extreme in
naturally-occurring distributions of starspots or faculae for young M
dwarfs \cite{Koen2021}.  Currently, the main argument against a
starspot-only explanation invokes the timescales and amplitudes of the
sharpest photometric features (CITE).  However, no independent
evidence has yet been presented for the presence of circumstellar
material in these objects.  Since transiting circumstellar
clumps would geometrically imply an occcurrence rate a few to ten
times the observed rate (CITE), the question of whether it is there
could potentially teach us about 10-30\% of M dwarfs during their
early lives.

@@

The dearth of evidence for circumstellar material around CPVs is
surprising given that separate studies of young BAFGKM stars have, for
decades, reported that stellar coronae contain both hot ($10^6$ K) and
cool ($10^4$ K) plasma. In particular, time-series spectroscopy has
shown periodic high-velocity absorption and emission in Balmer lines
such as H$\alpha$, caused by long-lived, corotating clumps of cool
plasma (CITE, CITE, CITE).  Such clumps are forced into corotation by
the magnetic field, and the exact geometry of where the plasma can
accumulate is dictated by the magnetic field's topology.
For instance, a tilted dipole field tends to yield an accumulation
surface of a warped torus \cite{Townsend2005}, whereas in the limit of
a single strong discrete field line, accumulation occurs along a fixed
point \cite{Waugh2022}.  However, none of these stars have shown any
photometric anomalies (CITE), leaving open the issue of whether these
two separate areas of study have any direct connection.  Nonetheless,
CPVs do respond to sudden magnetic field changes: there are many
documented cases of otherwise long-lived eclipse features disappearing
immediately following stellar flares (CITE, CITE).

@@

\subsection{Results}

@@

\subsection{Discussion}

Magetically-active, rapidly rotating stars with a wide range of masses
have been known to exhibit both sinusoidal emission features (CITE,
CITE, CITE, CITE) as well as sharp transient absorption features in
their line cores (CITE, CITE, CITE, CITE) similar to those in
Figure~\ref{fig:spec}.
None of the low mass stars showing such spectroscopic features have
previously shown abnormal light curves (CITE).
The preferred interpretation for such spectroscopic variability comes
from a loose analogy to quiescent solar prominences and filaments,
which are cool condensations of plasma in the solar corona that last
days to weeks (CITE).
This plasma is called a prominence when viewed in emission against the
dark backdrop of space, and a filament when viewed in absorption
against the solar disk.
In our Sun's magnetosphere, these condensations fall back to
the solar surface because gravity is stronger than any magnetic or
centrifugal force capable of sustaining them.
However for stars with magnetospheric radii $R_{\rm m}$ that exceed the
corotation radii $R_{\rm c}$, the effective potential experienced by
the plasma have local minima outside the $R_{\rm c}$, enabling the
material to be sustained for much longer timescales (see CITE, CITE).

Our Keck/HIRES observations were the first time a CPV was observed
with sufficient sensitivity to test for the presence of circumstellar
plasma clumps, and they showed for TIC~141146667 that they exist and
are corotating with the star.  
Characteristic densities and masses of these clumps are $n \sim
10^{10}$\,cm$^{-3}$ and $M \sim 10^{14}$\,kg (see~\ref{subsec:model}),
a similar density to solar prominences, but ten to one hundred times
more massive.
This observation rules out a ``starspot-only'' origin scenario
for CPVs, \cite{Koen2021} since such scenarios have no means of
explaining spectroscopic emission beyond the stellar disk.  Similarly,
scenarios in which the circumstellar material is made only of dust are
also ruled out.  Although dust may be present, to explain the
H$\alpha$ emission, the circumstellar clumps must include plasma with
a significant population of hydrogen atoms in the $n$=3 excited state.
While the plasma is undoubtedly sculpted by the star's magnetic field,
there are three a priori plausible origin locations for the material:
the star, an old and undetected disk, or outgassing rocky bodies.

The most intriguing potential known analogs to CPVs 
are the $\sigma$~Ori~E variables, a rare subset of B stars that almost
all have tilted-dipole magnetic fields (CITE) which can trap outflowing
stellar winds into warped plasma tori (CITE). 
These tori tend to have dense antipodal accumulations of plasma, and
the transits of these accumulations can produce broadband photometric
variability through a combination of bound-free scattering (CITE) and
Thomson scattering (CITE).  For $\sigma$~Ori~E
and most of its analogs, the result is light curves that almost all
appear ``simple'', resembling those of eclipsing binaries (CITE,
CITE).  The two known exceptions, HD~37776 and HD~64740, show complex
light curves resembling CPVs (Mikulasek2020, CITEBouma2024), and have
spectropolarimetric magnetic field maps indicating strong
contributions from higher-order magnetic moments (CITE CITE).  There are two
implications: firstly, the complexity of CPVs may be a direct
consequence of magnetic fields with highly multipolar contributions.
Secondly, CPVs could be a source of astrophysical false positives in
photometric searches for eclipsing binaries and transiting exoplanets
around young pre-main-sequence M dwarfs (e.g.  CITE).

% Need paragraph(s) on the future!
Pressing issues for future work on CPVs include determining the
composition and origin of the circumstellar material, understanding
the exact role of the stellar magnetic field, and exploring the
implied space weather experienced by the close-in rocky exoplanets
that, statistically, are likely to be nearby (CITE).

The composition -- either purely plasma, or else a dusty plasma -- can
be clarified by time-series optical and infrared spectrophotometry.
While observations of CPVs in the optical have previously suggested
the presence of dust (CITE), plasma clumps transiting over starspots
can also produce chromatic features (CITERackham).  The composition
and size distribution of any dust that is present could be most easily
resolved by measuring the extinction curve for one or more CPVs from
$\approx$1-10\,$\mu$m.  A composition similar to debris from rocky
bodies seen around white dwarfs (CITE Reach+09) would be indicative of
a rocky-body origin.  A composition closer to the ISM would be
indicative of condensed dust in an M dwarf wind, similar to that
formed in the environments of more evolved stars (CITE Marigo2008).

The role of the star's magnetic field could be better understood
through new observations, and new theory.
From the theoretical perspective, there is an urgent need for
rigid-field magnetohydrodynamic (MHD) modeling to go beyond previous work
(CITE CITE) by exploring the effects of non-dipolar field
contributions.
Fully time-dynamic MHD (e.g. CITE DaleyYates) will offer the
capability of understanding the connection between the plasma and the
dust.  %TODO: need this sentence to be better
Observationally, optical spectropolarimetry on sufficiently large
telescopes has the potential to assess both the field strength and
topology.
A more direct probe however might be to search for radio emission from
mechanisms such as the electron cyclotron maser instability, which can
provide a direct measurement of the field strength at the site of the
emitting region (typically an auroral oval).
Recent work (CITE) has shown that CPVs are variable radio emitters,
and that they show emission components that can be both persistent, as
well as short-lived and highly polarized.
%Observation in the x-ray would similarly inform our view of the hotter
%plasma in these magnetospheres which otherwise remains undetected.

It is currently unclear what, if any, relationship CPVs have to the
close-in exoplanets that exist around most M dwarfs (CITE).  However,
0.3-3\% of young M dwarfs show the CPV phenomenon (CITE),
and the circumstellar clumps of material need to be viewed
edge-on for the light curves to be ``complex''.  This implies that an
appreciable minority (3-30\%) of M dwarfs -- the rapidly rotating ones
with centrifugal magnetospheres -- seem to have similar circumstellar
environments to the CPVs.

@@

\subsection{Observations}

@@

\subsection{Data Reduction}

\subsection{Modeling the Emitting Clump}
\label{subsec:model}

>>>>

%\subsection{Introduction}
M dwarfs, stars with masses below about half that of the Sun, are the
only type of star to offer near-term prospects for detecting the
atmospheres of rocky exoplanets with water on their surfaces
\cite{NAP26141}.  Investment with JWST has proceeded accordingly, with
a few percent of the telescope's time so far having been allocated to
such objects.  In this context, it is therefore important to consider
how the evolution of an M dwarf might impact the evolution of its
planets.  Previous work has established that most M dwarfs host
close-in planets \cite{Dressing2015}, and that these planets are often
subject to long circumstellar disk lifetimes \cite{Ribas2015}, to
large doses of UV radiation \cite{France2013}, and to a high incidence
of flares and coronal mass ejections \cite{Gunther2020}.  However,
despite excellent work in these areas, the properties of the
circumstellar plasma and magnetospheric environments to which young,
close-in exoplanets are subject remain challenging to explore.

One glaring example of our current ignorance is the complex periodic
variables (CPVs).  Figure~\ref{fig:lc} highlights the main object of
interest in this article, but over one hundred analogous objects have
now been discovered by K2 and TESS 
\cite{Rebull2016,Stauffer2017,Rebull2018,Zhan2019,Rebull2020,Bouma2024}.
These CPVs are
defined by their highly structured and periodic optical light curves, 
and most are M dwarfs with rotation periods shorter than two days.
Within current sensitivity limits, none have primordial disks
\cite{Stauffer2017,Bouma2024}.
However, $\approx$3\% of stars a few million years old show this
behavior, and the observed fraction decreases to $\approx$0.3\%
by $\approx$150\,Myr \cite{Rebull2020}.

The two leading hypotheses to explain the CPVs are either that
transiting clumps of circumstellar material corotate with the star
\cite{Stauffer2017,Gunther2022,Bouma2024},
or that these stars represent an extreme in
naturally-occurring distributions of starspots or faculae for young M
dwarfs \cite{Koen2021}.  Currently, the main argument against a
starspot-only explanation invokes the timescales and amplitudes of the
sharpest photometric features.  However, no independent
evidence has yet been presented for the presence of circumstellar
material in these objects.  Since transiting circumstellar clumps
would geometrically imply an occcurrence rate a few to ten times the
observed rate, the question of whether it is there could potentially
teach us about 10-30\% of M dwarfs during their early lives.

@@

The dearth of evidence for circumstellar material around CPVs is
surprising given that separate studies of young BAFGKM stars have, for
decades, reported that stellar coronae contain both hot ($10^6$ K) and
cool ($10^4$ K) plasma. In particular, time-series spectroscopy has
shown periodic high-velocity absorption and emission in Balmer lines
such as H$\alpha$, caused by long-lived, corotating clumps of cool
plasma \cite{CollierCameron1989,Donati2000,Dunstone2006,Skelly2008}.
Such clumps are forced into corotation by the magnetic field, and the
exact geometry of where the plasma can accumulate is dictated by the
magnetic field's topology.  For instance, a tilted dipole field tends
to yield an accumulation surface of a warped torus
\cite{Townsend2005}, whereas in the limit of a single strong discrete
field line, accumulation occurs along a fixed point \cite{Waugh2022}.
However, none of these stars have shown any photometric anomalies
\cite{Bouma2024}, leaving open the issue of whether these two separate
areas of study have any direct connection.  Nonetheless, CPVs do
respond to sudden magnetic field changes: there are many documented
cases of otherwise long-lived eclipse features disappearing
immediately following stellar flares \cite{Stauffer2017,Bouma2024}.

@@

\section{Results}

@@

\section{Discussion}

Magetically-active, rapidly rotating stars with a wide range of masses
have been known to exhibit both sinusoidal emission features
\cite{Donati2000,Townsend2005,Dunstone2006,Skelly2008} as well as
sharp transient absorption features in their line cores
\cite{CollierCameron1989,CollierCameron1992,Cang2020} similar to those
in Figure~\ref{fig:spec}.  Such stars generally do not show complex
light curves \cite{Bouma2024}.  The preferred interpretation for their
spectroscopic variability comes from a loose analogy to quiescent
solar prominences and filaments, which are cool condensations of
plasma in the solar corona that can last days to weeks
\cite{VialEngvold2015}.  This plasma is called a prominence when
viewed in emission against the dark backdrop of space, and a filament
when viewed in absorption against the solar disk.  In our Sun's
magnetosphere, these condensations fall back to the solar surface
because gravity is stronger than any magnetic or centrifugal force
capable of sustaining them.  However for stars with magnetospheric
radii $R_{\rm m}$ that exceed their corotation radii $R_{\rm c}$, the
effective potential experienced by the plasma can have a local minimum
outside the $R_{\rm c}$, enabling the material to be sustained for
much longer timescales \cite{Petit2013,Daley-Yates2024}.
Generally speaking, such material need neither transit, nor be
optically thick.

Our Keck/HIRES observations are the first high-precision,
high-resolution time-series spectra of a CPV, and they show the
existence of circumstellar plasma clumps corotating with
TIC~141146667.  Characteristic densities and masses of these clumps
are $n \sim 10^{10}$\,cm$^{-3}$ and $M \sim 10^{14}$\,kg (see
Supplementary Methods Section~\ref{subsec:model}), a similar density
to solar prominences, but ten to one hundred times more massive.  This
observation rules out a ``starspot-only'' origin scenario for CPVs,
\cite{Koen2021} since such scenarios have no means of explaining
spectroscopic emission beyond the stellar disk.  Similarly, scenarios
in which the circumstellar material is made only of dust are also
ruled out.  Although dust may be present, to explain the H$\alpha$
emission, the circumstellar clumps must include plasma with a
significant population of hydrogen atoms in the $n$=3 excited state.
While the plasma is undoubtedly sculpted by the star's magnetic field,
there are three a priori plausible origin locations for the material:
the star, an old and undetected disk, or outgassing rocky bodies.

The closest potential known analogs to CPVs are the $\sigma$~Ori~E
variables, a rare subset of B stars that almost all have tilted-dipole
magnetic fields (CITE) which can trap outflowing stellar winds into
warped plasma tori \cite{Townsend2005,Townsend2008}.  These tori tend
to have dense antipodal accumulations of plasma, and the transits of
these accumulations can produce broadband photometric variability
through a combination of bound-free scattering \cite{Townsend2005} and
Thomson scattering \citep{Berry2022}.  For $\sigma$~Ori~E and most of
its analogs, the result is light curves that almost all appear
``simple'', resembling those of eclipsing binaries.  The two known
exceptions, HD~37776 and HD~64740, show complex light curves
resembling CPVs \cite{Mikulasek2020,Bouma2024} and have
spectropolarimetric magnetic field maps indicating strong
contributions from higher-order magnetic moments
\cite{Kochukhov2011,Shultz2018}.  There are two implications: firstly,
the complexity of CPVs may be a direct consequence of magnetic fields
with highly multipolar contributions.  Secondly, CPVs could be a
source of astrophysical false positives in photometric searches for
eclipsing binaries and transiting exoplanets around young
pre-main-sequence M dwarfs \cite{Johns-Krull2016,Bouma2020}.

Pressing issues for future work on CPVs include determining the
composition and origin of the circumstellar material, understanding
the exact role of the stellar magnetic field, and exploring the
implied space weather experienced by the close-in rocky exoplanets
that, statistically, are likely to be nearby \cite{Dressing2015}.

The composition -- either purely plasma, or else a dusty plasma -- can
be clarified by time-series optical and infrared spectrophotometry.
While observations of CPVs in the optical have previously suggested
the presence of dust \cite{Gunther2022,Koen2023}, a gray opacity
source such as electron scattering in a plasma transiting over
starspots could also produce chromatic features \cite{Rackham2018}.
The composition and size distribution of any dust that is present
could be most easily resolved by measuring the extinction curve for
one or more CPVs from $\approx$1-10\,$\mu$m.  A composition similar to
debris from rocky bodies seen around white dwarfs \cite{Reach2009}
would be indicative of a rocky-body origin.  A composition closer to
the ISM would be indicative of condensed dust in an M dwarf wind,
similar to that formed in the environments of more evolved stars
\cite{Marigo2008}.

The role of the star's magnetic field could be better understood
through new observations, and new theory.  From the theoretical
perspective, there is an urgent need for rigid-field
(magneto)-hydrodynamic modeling to go beyond previous work
\cite{Townsend2005,Townsend2008} by exploring the effects of
non-dipolar field contributions.
Fully time-dynamic MHD \cite{Daley-Yates2024} will offer the
capability of understanding the connection between the plasma and the
dust.  %TODO: need this sentence to be better
Observationally, optical spectropolarimetry on sufficiently large
telescopes has the potential to assess both the field strength and
topology.  A more direct probe however might be to search for radio
emission from mechanisms such as the electron cyclotron maser
instability, which can provide a direct measurement of the field
strength at the site of the emitting region (typically an auroral
oval).  Recent work \cite{Kaur2024} has shown that CPVs are variable
radio emitters, and that they show emission components that can be
both persistent, as well as short-lived and highly polarized.
%Observation in the x-ray would similarly inform our view of the hotter
%plasma in these magnetospheres which otherwise remains undetected.

It is currently unclear what, if any, relationship CPVs have to the
close-in exoplanets that exist around most M dwarfs
\cite{Dressing2015}.  However, 0.3-3\% of young M dwarfs show the CPV
phenomenon \cite{Rebull2020}, and the light curves being ``complex''
seems to be associated with the circumstellar clumps of material
transiting the star.  This geometric requirement implies that an
appreciable minority (3-30\%) of M dwarfs -- the rapidly rotating ones
with centrifugal magnetospheres -- have similar circumstellar
environments to the CPVs.

@@

\section{Observations}

@@

\section{Data Reduction}

\section{Modeling the Emitting Clump}
\label{subsec:model}
